---
title: "PCA"
author: "Lucas Mendicino"
date: "11/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggcorrplot)
library(FactoMineR)
```

```{r}
summary(mtcars)
```

```{r}
correl <- cor(mtcars, use = "complete.obs")
```

```{r}
ggcorrplot_clustered <- ggcorrplot(correl, hc.order = TRUE, type = "lower")
ggcorrplot_clustered
```

```{r}
mtcars_pca <- PCA(mtcars)
```
Factor map: first PC
-  cyl, disp, wt take positive values
- mpg takes negative values
1st PC separates cars that are powerful and heavy on the right side and cars that are economical on the left side
2nd PC: speed
- cars with lots of gears are placed higher
- qsec takes lower values

```{r}
summary(mtcars_pca)
```


```{r}
# Get the variance of the first 3 new dimensions.
mtcars_pca$eig[,2][1:3]

# Get the cumulative variance.
mtcars_pca$eig[,3][1:3]

```


```{r}
dimdesc(mtcars_pca, axes = 1:2)

```

```{r}

# Trace variable contributions in mtcars_pca
mtcars_pca$var$contrib
```

```{r}
library(ade4)
```

```{r}
mtcars_pca2 <- dudi.pca(mtcars, scannf = FALSE)

summary(mtcars_pca2)
```

plotting contributions

```{r}
library(factoextra)
fviz_pca_var(mtcars_pca,
             col.var = "contrib",
             gradient.cols = c("#bb2e00", "#002bbb"),
             repel = TRUE)

```

contributions of selected variables
```{r}
fviz_pca_var(mtcars_pca,
             select.var = list(contrib = 4),
             repel = TRUE)

```

```{r}
fviz_contrib(mtcars_pca,
             choice = "var",
             axes = 1,
             top = 5)

```

how accurate the representation of our cariables on the PC plane are
```{r}
fviz_pca_ind(mtcars_pca,
             col.ind = "cos2",
             gradient.cols = c("#bb2e00", "#002bbb"),
             repel = TRUE)

```
barrploting cos2 for individuals
```{r}
fviz_cos2(mtcars_pca,
             choice = "ind",
             axes = 1,
             top = 10)

```
biplots

```{r}
fviz_pca_biplot(mtcars_pca)

```

adding ellipsoids

```{r}
mtcars$cyl <- as.factor(mtcars$cyl)

fviz_pca_ind(mtcars_pca,
             label = "var",
             habillage = mtcars$cyl,
             addEllipses = TRUE)

```











































